---
title: "Data wrangling with R - P1 Renaming"
author: "Sumy V Baby, Tomasz Zielinski"
date: "2022-03-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Welcome to data wrangling with R.

This and the following R-Markdown notebooks have been inspired by our real data curation work for the project about monitoring COVID virus levels in wastewater around Scotland ([see project page](https://biordm.github.io/COVID-Wastewater-Scotland/)).

Firstly, check our raw data.

## Inspecting RAW DATA

Our raw data are included in the raw_data folder of this project. Let's check what it contains

```{r}
list.files("../raw_data")
```
*** The `../raw_data` means that we are going to navigate to the parent folder (we are in src folder) cause we use `..` and then we will list the content of the `raw_data` folder. ***

Let's read measurements from the COVID RNA Monitoring Project file

```{r}
raw_data <- read.csv("../raw_data/COVID_RNA_Monitoring.csv")
head(raw_data)

```
`read.csv` read content of the file into `raw_data` variable.  
`head(raw_data)` printed first few rows of the data.

To have it as a nice wide table rather than wrapped around when generating r-markdown report we can use:

```{r}
knitr::kable(
  head(raw_data),
  caption = "COVID RNA levels in wastewater"
)
```
Some of the columns have . in their name, it is because the original file contained columns names with space. It is generally a bad practice to have spaces in columns names.
Also some columns are not descriptive enough, for example `HBOARD` stands for health board and the `Date` is actually a date on which the samples were collected.

So let's rename the columns.

## Renaming columns

```{r}
renamed_data <- raw_data
names(renamed_data)[1] <- "Health_Board"
names(renamed_data)[3] <- "Date_collected"
head(renamed_data[1:5])

```
`names` in base R can be used to both read the column names or to set it, depending if it is used in assignment or not.

That approach is error prone and tedious as we need to use the correct column index (see that date was in column 3 not 2).

### Renaming by selection
It would be better if we could select which columns we want to rename and provide new names.  
For example:

```{r}
old_name = c("HBNAME","Date","Date.Analysed")
new_name = c("Health_Board","Date_collected","Date_analysed")
renamed_data <- raw_data
names(renamed_data)[names(renamed_data) %in% old_name] = new_name
head(renamed_data[1:5])

```
We still need to provide the new names in the right order!

Probably less error prone is actually to get all the existing names, change them and set back.

### Renaming all
For that we need a vector with new names, we could generate a text template with a following command
```{r }
print(paste0("'", names(raw_data),"'", collapse=","))
```
We used ' so that the escape \" character won't be printed
Now we can edit the columns ones by one and use to set the new names
```{r}
new_name = c('Health_Board','Site','Date_collected','Date_analysed','SW_sample_number','N1.Description','N1.Reported.Value','N1.Sample.1','N1.Sample.2','N1.Sample.3','Calculated.Mean','Standard.Deviation','Flow..l.day.','Ammonia..mg.l.','pH.Value','Modelled.Flow..l.day.','Million.Gene.Copies.Per.Person.per.Day','Analysis.Lab','Row_Count')
renamed_data <- raw_data
names(renamed_data) = new_name
head(renamed_data[1:5])

```

The best solution would be to provide names dictionary, which will match old names with new names. That way we will be resilient to changes in the columns order.

### Renaming with dictionary

```{r}
dictionary <- data.frame(
  old_name = c("Date",          "Date.Analysed", "HBNAME"),
  new_name = c("Date_collected","Date_analysed", "Health_Board")
)
renamed_data <- raw_data
names(renamed_data)[match(dictionary$old_name, names(renamed_data))] <- dictionary$new_name
head(renamed_data[1:5])
```
Look how the first column was renamed even it was the last one in our dictionary.

We could of course create a file with the old and new names, and read the mappings from there. Try to do it yourself.

So lets rename all the columns using two lists of names


```{r}
dictionary <- data.frame(
  old_name = c('HBNAME','Site','Date','Date.Analysed','SW.Sample.Number','N1.Description','N1.Reported.Value','N1.Sample.1','N1.Sample.2','N1.Sample.3','Calculated.Mean','Standard.Deviation','Flow..l.day.','Ammonia..mg.l.','pH.Value','Modelled.Flow..l.day.','Million.Gene.Copies.Per.Person.per.Day','Analysis.Lab','X.Row.Count.'),

  new_name = c("Health_Board","Site","Date_collected","Date_analysed", "SW_sample_number","N1_description", "N1_Reported_value-gc_per_L","N1_Repl_1-gc_per_L","N1_Repl_2-gc_per_L","N1_Repl_3-gc_per_L","Calculated_mean","Standard_Deviation","Flow-L_per_day","Ammonia-mg_per_L","pH_value","Modelled_flow-L_per_day","Million_gene_copies_per_person_per_day","Analysis_lab","Row_Count")
)
renamed_data <- raw_data
names(renamed_data)[match(dictionary$old_name, names(renamed_data))] <- dictionary$new_name
head(renamed_data[1:5])
```

Let's save our renamed data into a file.

```{r}
if (!file.exists('../data')) {
  dir.create(file.path('..', 'data'))  
}

fName = '../data/renamed_data.csv'
write.csv(renamed_data, fName, row.names = FALSE)

```

We first checked if the output directory exists, if not create it and then we saved the file.

### Automatic correction

You may noticed that in majority of the cases we just replaced . with _ in our new columm names.
Segure it could be done in some automatic way.

Indeed, it can be done using pattern matching with regular expressions. 
You will need to read more about regular expressions on your own, but here we will deal with replacing one character with other.

The command to replace is called sub or gsub (the first one replace only the first appearance) and we pass it to it "what to replace", "with what" and "Where". 

So lets try it
```{r}

sub(".","_",names(raw_data))
```

OK that does not look good, it is cause '.' means any character, so every starting letter was replaced with _. We need to "escape" the  '.', using \\ which says it is just a . character

```{r}

sub("\\.","_",names(raw_data))
```

That worked, but only for the first ., we need gsub

```{r}

gsub("\\.","_",names(raw_data))
```

If we look at 13, 'Flow__l_day_' looks weird it is cause it was 'Flow..l.day.' with two dots.
We can tell gsup to match multiple consecutive appearence of a character with + (which means at least 1), or * which means (0 or many)

```{r}

gsub("\\.+","_",names(raw_data))
```
Now the only annoying part is the last, handing _

We can match the end of word with $ sign, so we can replace only the last occurence.
For example,

```{r}

gsub("\\.$","ZZZ",names(raw_data))
```
Attached ZZZ to the last ., what we want is replacing the last . with empty string ""


```{r}

gsub("\\.$","",names(raw_data))
```
We can chain our two operations (it is called pipe) by using `%>%`

```{r}
library(magrittr)
gsub("\\.$","",names(raw_data)) %>% gsub("\\.+","_",.)
```
The first gsub removed the last ., the second replaced other .(.) with _
Look like in the second gsub we passed the "Where" using . as argument, it means that the
output of first gsub was passed to the second. 

For this code to work we needed to loead library which defines the pipe %>% library(magrittr).

We can do it for better readibility also for the first gsub and finally use it set the column names

```{r}
names(raw_data) %>%
    gsub("\\.$","", .) %>% 
    gsub("\\.+","_",.) -> new_name

new_name[1] <- "Health_Board"
new_name[3] <- "Date_collected"

renamed_data = raw_data
names(renamed_data) = new_name
head(renamed_data[1:5])
```

It is not exactly the same as our manually curated names, but it could be good enough for your use cases.