---
title: "Data wrangling with R - P2 Sampling sites file generation"
author: "Sumy V Baby, Tomasz Zielinski"
date: "2022-03-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We continue our work with the data, now we will focus on Sampling sites file generation

This and the following R-Markdown notebooks have been inspired by our real data curation work for the project about monitoring COVID virus levels in waste water around Scotland ([see project page](https://biordm.github.io/COVID-Wastewater-Scotland/)).

## Reading the data files after loading necessary libraries

We need to read the curated_site_coordinate_data.csv and curated_population_data.csv files first

```{r}
library("lubridate")
library("sgo")
library("tidyverse")

site_coordinates<- read.csv("../data/curated_site_coordinate_data.csv")
site_population<- read.csv("../data/curated_population_data.csv")
glimpse(site_coordinates)
glimpse(site_population)

```
We can use command `glimpse` to see all the columns (variables) and first data in each column.

As we can see the Latitude and Longitude measurements are in different system (UK grid). Now we need to convert geographic locations into GPS DD unit. 
The function "sgo_points" in "sgo" library take the "Longitude" and "Latitude" from site_coordinates which is in UK grid system (specified by espg: 27700) and converts to an object of class sgo_points.

```{r}
locs <- sgo_points(site_coordinates, coords=c("Longitude", "Latitude"), epsg=27700)

```

Now we can use the function  "sgo_bng_lonlat" from "sgo" library that converts Ordnance Survey grid reference easting/northing coordinates to GCS longitude/latitude. The "to=4326" in the function specifies epsg code. EPSG Geodetic Parameter Dataset (also EPSG registry) is a public registry of geodetic datums, spatial reference systems, Earth ellipsoids, coordinate transformations and related units of measurement. EPSG:4326 - WGS 84, latitude/longitude coordinate system based on the Earth's center of mass, used by the Global Positioning System among others.

```{r}
Data_sgo_4326 <- sgo_bng_lonlat(locs, to=4326)
Data_4326 <- as.data.frame(Data_sgo_4326)
head(Data_4326)
```

We can see that the Column names needs to be changed. x here means Longitude_dd and y means Latitude_dd. Let's change that.

```{r}
colnames(Data_4326) = c("Longitude_dd", "Latitude_dd", "Health_Board","Site")
head(Data_4326)
```

Now let's merge the coordinate information with population data and check how it looks

```{r}
merged <- merge( site_population, Data_4326,by = "Site")
head(merged)
```

As we know that there are Health_Board information in both site_population and Data_4326. So when these are merged we can see two columns Health_Board.x from site_population and Health_Board.y from Data_4326. We need to select one of these columns as both are the same. We can avoid the column 3 that is "Health_Board" in Data_4326 while merging and check how it looks.

```{r}
merged <- merge( site_population, Data_4326[,c(1,2,4)],by = "Site")
head(merged)
```
Now we need to arrange the columns in specific order as : Site, Health_Board, Latitude_dd, Longitude_dd, Population, Population_Band. We can do it by simply rearranging by column numbers or using the "select" function 
```{r}
merged <- merged[,c(1,2,6,5,4,3)]
#merged <- select(merged,Site,Health_Board,Latitude_dd,Longitude_dd,Population,Population_Band) 
head(merged)
```

The file is ready. We can create the out directory if it doesn't exist and save the file

```{r}
if (!file.exists('../out')) {
  dir.create(file.path('..', 'out'))  
}

fName = '../out/sampling_sites.csv'
write.csv(merged, fName, row.names = FALSE)
```

