---
title: "Data wrangling with R - P2 Full data file generation"
author: "Sumy V Baby, Tomasz Zielinski"
date: "2022-03-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We continue our work with the data, now we will focus on Sampling sites file generation

This and the following R-Markdown notebooks have been inspired by our real data curation work for the project about monitoring COVID virus levels in waste water around Scotland ([see project page](https://biordm.github.io/COVID-Wastewater-Scotland/)).


## Reading the data files after loading necessary libraries

We need to read the curated_site_coordinate_data.csv and curated_population_data.csv files first. As we know that the curated_data.csv contains column names having "-", the option "check.names = F" is used to read it as it is. Otherwise the the hyphen is converted in to dot symbol while reading the file.

```{r}
library("lubridate")
library("tidyverse")

curated_data<- read.csv("../data/curated_data.csv", check.names = F)
sampling_sites<- read.csv("../out/sampling_sites.csv")
head(curated_data)
head(sampling_sites)

```

Now lets merge the two data sets by site. Lets avoid the "Health_Board" from the curated_data by not selecting column 2.

```{r}
merged<-merge(sampling_sites[,c(1,3,4,5,6)],curated_data,  by = "Site")
head(merged)
```

We should arrange the columns in the order : 
Health_Board, Site, Latitude_dd, Longitude_dd,Population,Population_Band,Date_collected,Date_analysed,SW_sample_number,N1_description,"N1_Reported_value-gc_per_L","N1_Repl_1-gc_per_L","N1_Repl_2-gc_per_L","N1_Repl_3-gc_per_L","Calculated_mean","Standard_Deviation","Flow-L_per_day","Ammonia-mg_per_L","pH_value","Modelled_flow-L_per_day","Million_gene_copies_per_person_per_day","Analysis_lab".
We can drop the last column:23 as it is not of much significance. Another way of doing it is using "select" function. The column names containing "-" should be kept in double inverted commas in this case. otherwise R will not read it properly. 


```{r}
merged_full<- merged[c(6,1,2,3,4,5,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22)]

#merged_full<-select(merged,Health_Board,Site,Latitude_dd,Longitude_dd,Population,Population_Band,Date_collected,Date_analysed,SW_sample_number,N1_description,"N1_Reported_value-gc_per_L","N1_Repl_1-gc_per_L","N1_Repl_2-gc_per_L","N1_Repl_3-gc_per_L","Calculated_mean","Standard_Deviation","Flow-L_per_day","Ammonia-mg_per_L","pH_value","Modelled_flow-L_per_day","Million_gene_copies_per_person_per_day","Analysis_lab")

head(merged_full)
```

Now the file is ready. We can create the out directory if it doesn't exist and save the file

```{r}
if (!file.exists('../out')) {
  dir.create(file.path('..', 'out'))  
}

fName = '../out/data_full.csv'
write.csv(merged_full, fName, row.names = FALSE)
```

