---
title: "Data wrangling with R - P2 HeatMap Generation-Determining breakpoints "
author: "Sumy V Baby, Tomasz Zielinski"
date: "2022-03-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
This code calculate break points to be used in the heatmap color scale. Although the raw gene copies per L data has clear technical thresholds for positive data,normalization creates overlap between "Weak positive", "Positive DNQ" and "Positives". The results in each group have long tails so min/max was not an option. For that reason 3rdquartiles are used to mark start of the positive data and to truncate the color scale for better visibility


This and the following R-Markdown notebooks have been inspired by our real data curation work for the project about monitoring COVID virus levels in waste water around Scotland ([see project page](https://biordm.github.io/COVID-Wastewater-Scotland/)).


## Reading the data files after loading necessary libraries

We need to read the data_full.csv first

```{r}
library("tidyverse")

allData <- read.csv("../out/data_full.csv", check.names = F)

```
We can take filter out the positive normalized values and non-positive normalized values for determining the breakpoints.

The "filter" function from "tidyverse" library filters "allData" dataframe for "Positive"  N1_description values. Then the  corresponding values for column 
"Million_gene_copies_per_person_per_day" are assigned to "allPositive".

Similarly "filter" function is used to filter "allData" dataframe for all non positive N1_description values. Then the  corresponding values for column 
"Million_gene_copies_per_person_per_day" are assigned to "nonPositive".

```{r}
allPositive <- filter(allData, allData$N1_description == "Positive")$Million_gene_copies_per_person_per_day
nonPositive <- filter(allData, allData$N1_description %in% c("Weak Positive","Positive (DNQ)","Negative"))$Million_gene_copies_per_person_per_day
```

Let's now take 75th percentile value for the positive and non-positive normalized data

```{r}

pos_start = round(quantile(nonPositive, 0.75, na.rm=TRUE), digits = 1)
pos_end = round(quantile(allPositive, 0.75, na.rm=TRUE), digits = 1)

print(pos_start)
print(pos_end)

```



```

